Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

@@

We define a \textbf{program order pomset} to be a pomset with \(\mc{A} =
\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
program order pomsets given by sequential composition for \(\mc{A}\). This
yields an \isotopessa model with support for concurrent read and write
operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
semantics
\begin{equation}
  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
  \qquad
  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
\end{equation}
The semantics of \(\ms{read}\) in particular give a hint as to how this model
works: rather than tracking the state of the heap, we simply \textit{emit} a
pomset of the events that would have to happen for a given execution (in the
case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
the case of a write, the single write event \(x := v\) for a write of \(v\)),
and later post-filter on the pomset of an entire program to get a sequentially
consistent execution. Hence, our semantics remains compositional, allowing us to
reason about each program fragment individually, while still allowing us to add
arbitrary external state (e.g. another program concurrently modifying the heap)
before finally considering the execution of an entire program.

@@

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}_w^*\)
of lists of write actions, which we will interpret as linear pomsets ordered by
index. In particular, we define the monad \(\ms{TSO} =
\ms{StateT}\;\ms{Buf}\;\ms{PO}\). Given a buffer \(\ms{Buf}\), we can define the
result of reads \([\cdot]_x: \ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the
buffer as follows:
\begin{equation}
  (L;\{x := v\})[x] = v
  \qquad
  (L;\{y := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! We will implement TSO ordering by buffering all
our writes, in which case they are only visible to the local thread. At any
point, we may choose to \textit{flush} some of the writes in buffer to the
global pomset of events. This is implemented via the morphism (called
\(\ms{split}\) in \cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), L', L'') | L = L';L''\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
On the other hand, reads events will first attempt to read from the buffer, and,
if there is no corresponding write in the buffer, will read an arbitrary value,
in both cases pushing an event to the global pomset; that is, the semantics of
reads and writes are given by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, L)\});\ms{pflush}
  \end{aligned}
\end{equation}

\TODO{new read/write}

\TODO{in the presence of parallel composition}

>>>>

% Weak memory
\newcommand{\bufloc}[1]{\overline{#1}}

@@

We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}} =
\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
program order pomsets given by sequential composition for \(\mc{A}\). This
yields an \isotopessa model with support for concurrent read and write
operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
semantics
\begin{equation}
  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
  \qquad
  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
\end{equation}
The semantics of \(\ms{read}\) in particular give a hint as to how this model
works: rather than tracking the state of the heap, we simply \textit{emit} a
pomset of the events that would have to happen for a given execution (in the
case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
the case of a write, the single write event \(x := v\) for a write of \(v\)),
and later post-filter on the pomset of an entire program to get a sequentially
consistent execution. Hence, our semantics remains compositional, allowing us to
reason about each program fragment individually, while still allowing us to add
arbitrary external state (e.g. another program concurrently modifying the heap)
before finally considering the execution of an entire program.

@@

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we will to introduce a \textit{buffer} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. At any point, we may choose to \textit{flush} some
of the writes in buffer to the global pomset of events.
\TODO{\(\mc{A}_{\ms{TSO}}\), \(\mc{A}_b\)}
of type \(\ms{Buf} = \mc{A}_b^*\)
of lists of write actions, which we will interpret as linear pomsets ordered by
index. In particular, we define the monad \(\ms{TSO} =
\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\). 
\TODO{using \(\mc{A}_{\ms{TSO}}\)}
Given a buffer \(\ms{Buf}\), we can define the
result of reads \([\cdot]_x: \ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the
buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! 
\TODO{segue}
This is implemented via the morphism (called
\(\ms{split}\) in \cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
\TODO{reflow}
On the other hand, reads events will first attempt to read from the buffer, and,
if there is no corresponding write in the buffer, will read an arbitrary value,
in both cases pushing an event to the global pomset; that is, the semantics of
reads and writes are given by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
\TODO{text explaining this}
\TODO{in the presence of parallel composition}
